I/O simulation model with Linux page cache, integration and evaluation in SimGrid framework

### 2020-11-16 paper_io_sim.tex
Draft version of the paper

\author{\IEEEauthorblockN{Hoang-Dung Do, Val\'erie Hayot-Sasson, Rafael Ferreira da Silva, Henri Casanova, Tristan Glatard
  }\\
  \IEEEauthorblockA{
    Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada
  }
}

@@

	\begin{abstract}
		\begin{itemize}
			\item The I/O bottleneck in HPC and the need of experiments.
			\item HPC experiment frameworks and advantages of SimGrid.
			\item The missing of the ability to simulate page cache, the goal of this paper.
			\item Principle of the simulator, experiment scenarios and comparisons.
			\item Brief discussion on results and future work.
		\end{itemize}
	\end{abstract}

	\section{Introduction}
		\begin{itemize}
			\item HPC, the bottleneck in I/O and the demand of HPC experiments. 
			\item Difficulties in conducting high performance computing experiments and the need of simulation frameworks.
			\item Existing experiment methods, simulators, simulation frameworks. The advantages of SimGrid compared to others \cite{simgrid2008, lebre2015}. The missing of the ability to simulate page cache in SimGrid \cite{lebre2015}.
			\item The objective of the paper: Add capability to simulate I/O with page cache in SimGrid.
			\item Clarify, distinguish between memory I/O (i/o to page cache) and disk I/O.
		\end{itemize}
	\section{Related Work}

		\subsection{Page cache}
			\begin{itemize}
				\item What is page cache? How it works \cite{linuxdev3rd2010}. 
				Write-back vs writethrough and direct i/o. Effects and importance of page cache.
				\item Introduce some existing strategies with some highlighted pros and cons.
				\item Current implementation in Linux and some reasons why it is chosen to be implemented (implementation complexity, effectiveness, overhead, etc) \cite{linuxdev3rd2010}
				\item \tristan{LRU lists are mentioned in the next section, they should be described here.}
			\end{itemize}

		\subsection{Simulators}
			\begin{itemize}
				\item Discuss some existing methods, simulation frameworks to conduct HPC experiments. Compare pros and cons (accuracy, simulation time, usability) of some simulators (SimGrid, GridSim).
				\item Related development: RAM energy consumption \cite{gill2019} \cite{ouarnoughi2017} 
				\item What is implemented in simulators that useful for our model:
				distinguished and shared bandwidths, concurrency, device latency. 
				\item Discuss the pros of SimGrid and the reasons why we chose it to conduct experiments. (Section 2.2.2 in \cite{simgrid2014})
				\item Introduce WRENCH and it's advantage.
				\item \tristan{Mention the status of storage simulation in simulators, so that we understand from Fig 1 that storage devices are available}
			\end{itemize}

	\section{Method}

@@

			We leverage existing storage simulation models to simulate the memory,
			which is characterized with storage capacity, distinguished read/write 
			bandwidths and device latency. 
			All of these features are dealt within models in simulation 
			frameworks, which makes storage models applicable to memory \cite{lebre2015}. 
			In addition, leveraging existing disk models allows us 
			to take advantage of bandwidth sharing between concurrent disk I/O to 
			simulate parallel memory accesses.\tristan{we should refer to related work, 
            see comment there} \valerie{Agreed. It's unclear how existing storage models enable this at the moment}.

@@

			When simulated applications need to read/write a file, they can simply call 
			the IO Controller instead of the storage simulation models. 
			\tristan{this should move, not sure where for now: 
			Because every computer has its own memory and system configurations, 
			one IO Controller object and one Memory Manager object are created 
			for each simulated host.}

			As mentioned, in our model, a file read/write can be complete by read/write 
			chunk-by-chunk until entire file is read/written.
			To read or write a chunk of file, a simulated application sends a request to the 
			IOController instance on the node where the requested file is stored.
			Based on the memory status in Memory Manager, the IOController 
			orchestrates cache and disk reads/writes, flushing, cache eviction 
			to simulate I/O time.

@@

			We use our dedicated cluster hosted at Concordia University to conduct 
			the experiments. The cluster consists of one login node, 8 compute nodes 
			and 4 storage nodes connected with two network switches. The login node 
			has one Intel(R) Xeon(R) Gold 6130 CPU @ 2.10GHz, 137~GB (128~GiB) of RAM, 
			1.8TB of storage of XFS file system, 13~GB of tmpfs file system and 385 TB of 
			Lustre shared file system. Each compute node has two 16 cores Intel(R) 
			Xeon(R) Gold 6130 CPU @ 2.10GHz, 275~GB (256~GiB) of RAM and 6 SSDs, 
			450~GB each with XFS file system, 378~GB of tmpfs and 126~GB of devtmpfs file system.
			Lustre file system is configured with one metadata target with 854~GB 
			of storage, 44 object storage targets with 8.7~TB of storage each. 
			The cluster is run on CentOS 8.1 with the Slurm Workload Manager installed. 
			We use \textbf{\textit{atop}} and \textbf{\textit{collectl}} as tools to 
			monitor and collect memory status and disk throughput 
			on the cluster. 
			\textcolor{red}{[Add network description here, include number of disks 
			on all nodes ,maybe add a diagram]}

@@

			\subsubsection{Evaluation on remote storage}

		    Since remote storage is common in HPC, we conduct another experiment with 
		    the same scenario as in the multi-threaded experiment, except that files 
		    are stored on a node and read by another node.

@@

	\section{Results}

		\begin{itemize}

			\item Quantitative results: 
				\begin{itemize}
					\item Relative errors of simulation time and memory used compared to real results.
					\item Simulation time compared to SimGrid and original WRENCH.
				\end{itemize}

			\item Ability of the model to generalize memory trends (dirty data, cache used) and disk throughput.

			\end{itemize}

		The benchmark results of bandwidths and the values used in the simulators are shown in \ref{table:benchmark}. These configurations are used in all experiments.

		\begin{table}
		\centering
		\begin{tabular}{|l|c|c|c|}
		\hline
			Bandwidths (MBps)  & Cluster (real) & Python simulator & WRENCH simulator\\
		\hline
			Memory read  & 6860	& 4812	 & 4812\\
			Memory write & 2764	& 4812 & 4812\\
			Disk read & ~510 & 465 & 465\\
			Disk write & ~420 & 465	 & 465\\
		\hline
		\end{tabular}
		\caption{Bandwidth benchmark results and simulator configurations}
		\label{table:benchmark}
		\end{table}

@@

			The I/O time of tasks in simulators and real execution is also tracked and compared. 
			Figure~\ref{fig:single_errors} shows the relative error of simulated I/O time 
			of individual I/O activities. 
			Our pipeline consists of 3 tasks, and each task reads an input file, 
			performs some computation, and writes output file. 
			Thus, there are 6 sequential I/O activities, \textit{read1} and \textit{write1} 
			from the first task, \textit{read2} and \textit{write2} the second task, 
			and \textit{read3}, \textit{write3} from the last task.

			In general, with all input sizes in the experiment, the simulation errors 
			of simulators with page cache model (Python and WRENCH with page cache simulators) 
			are remarkably lower than the result with original WRENCH . 
			The reason is, in original WRENCH, because page cache is not simulated, 
			data is only read/written with disk bandwidth, which is much slower 
			than memory bandwidths, leading to simulated I/O time longer than in reality.
			Especially, the simulation time errors of \textit{read3} in the simulators 
			with page cache are just under the error in original WRENCH.

@@

			\begin{figure*}		
			\begin{subfigure}{\linewidth}
				\centering
   				\includegraphics[width=\linewidth]{result/multi/figures/multi_local.pdf}
			\end{subfigure}		
			\caption{I/O time with different numbers of concurrent pipeline}
			\label{fig:multi_local}
			\end{figure*}

@@

@@

	\section{Discussion and Future Work}
		\begin{itemize}
			\item Read order: disk, inactive list, then active list. 
			\item Cache eviction mechanism not evicting active list, leads to evicting file being written.
			\item Sensitivity of the simulator on the variation of memory and disk bandwidth. 
			\item Our model does not support random I/O and readahead.
			\item Disk state save/restore when hosts are turned on/off.
		\end{itemize}
	\section{CFI for cluster}
\bibliographystyle{plain}
\bibliography{citation}

>>>>

\author{\IEEEauthorblockN{Hoang-Dung Do$^1$, Val\'erie Hayot-Sasson$^1$, Rafael Ferreira da Silva$^3$, Henri Casanova$^2$, Tristan Glatard$^1$
  }\\
  \IEEEauthorblockA{
    $^1$Department of Computer Science and Software Engineering, Concordia University, Montreal, Canada \\
     $^2$Department of Information and Computer Sciences, University of Hawai‘i at Manoa, U.S.A \\
     $^3$Department of Computer Science, University of Southern California, Los Angeles, CA, USA
  }
}

@@

	\begin{abstract}

		The data explosion era seen in recent years results in growing needs in 
		data storing, data analysis. 
		Large computing systems including grids, clusters, clouds, peer-to-peer, 
		volunteer computers have been increasingly used to solve the problem. 
		Despite of their powerful computing capability and technological improvements, 
		there are still bottlenecks in these systems. 
		When it comes to I/O, a bottleneck can be seen in the disk I/O due to hardware 
		limitations. Data caching using page cache is one of the solutions applied 
		to minimize disk accesses and improve the I/O performance. 
		In computing system and application optimization studies, one of the 
		effective approaches is using simulation tools to conduct experiments instead 
		of running applications on real infrastructures. 
		Although simulators are built with individual simulation model for each computing 
		resource, the majority does not take page cache into consideration, 
		which can lead to low simulation accuracy in many cases.

		In this paper, we propose a simulation model page cache to simulate I/O 
		with key features of page cache. 
		The model is build with the consideration to allow it to be integrated to 
		existing simulation tools. 
		We also implement the model with SimGrid using WRENCH. 
		Then we create some simulators for different use cases. 
		Finally, we analyze the results, compare them to results of the existing 
		tool as well as real execution.

	\end{abstract}

	\section{Introduction}

		In recent years, as a result of the growth of data generated from a variety 
		of data sources, we have seen increasingly growing demands in data storage, 
		processing and analysis. 
		Storing and processing such a massive amount of data require large 
		computing infrastructures. 
		Data processing is a complex process whose the performance is affected 
		by many factors and hindered by bottlenecks in the systems.
		One of the bottlenecks that should be taken into consideration is disk access 
		time due to hardware limitations. 
		Beside hardware technical improvements, architectural solutions have been 
		proposed and implemented in computing systems in order to tackle the 
		disk I/O problem. 
		Caching, a successful technique applied into memory to maximize the 
		I/O performance by keeping data in memory for future accesses,
		has been proving the effectiveness when employed in computer systems.

		Studying in optimizing not only the utilization of page cache but also other 
		computing resources is a need that has never been overlooked.
		The research on computing infrastructures to maximize performance, 
		shorten processing time, and minimize costs gets involved with experiments. 
		However, real platforms are complex and not built for the experimental 
		purposes due to limited and shared resources, operational costs, unstable 
		environments and the result reproducibility difficulties. 
		An alternative approach is to used simulation tools developed for research 
		and experimental purposes.
		The obvious advantages of using simulation are the freedom of choosing 
		simulated infrastructures, reproducible results and low costs.
		These simulators are developed and integrated with simulation models 
		of computing resources including CPU, network and storage.
		However, some reviews over many simulators report that there has been 
		a trade-off between accuracy and scalability due to the level of details in 
		simulation models \cite{casanova2014simgrid} \cite{byrne2017review}.
		Moreover, different simulators are developed by and for a specific community,
		which hinders them from being used in cross-domain.
		SimGrid is a simulation framework developed with the objectives of  
		balancing between accuracy and scalability, as well as versatility.

		In spite of the importance of page cache and the impacts it has on I/O 
		operations, its simulation model is absent in many simulators. 
		Some simulation tools try to model the page cache with cached data, 
		but their models lack of some key features \cite{nunez2012simcan} 
		\cite{nunez2012icancloud}. 
		Moreover, these simulators have a problem in scalability with 
		long simulation time regarding to their very detailed level of simulation.
		Some other works are able to capture some important features of page cache, 
		but they focus on very specific domains \cite{xu2018saving}.

		In this paper, we propose a simulation model of I/O with page cache.
		The model is integrated in WRENCH, an open-source library built 
		on top of SimGrid with more abstract API. 
		Then we evaluate our model with four simulators in four 
		experiment scenarios. 
		This paper is organized as follows. 
		In Section \ref{relatedwork} we provide details of page cache 
		existing simulation tools.
		In Section \ref{relatedwork} we propose our I/O with page 
		cache simulation model and how we implement it and our experiments 
		to evaluate the model.  
		Section \ref{results} compares the results of our model with the results 
		of original WRENCH and real execution results. 
		Finally, Section \ref{discussion} concludes and presents future 
		work.

	\section{Related Work}		
	\label{relatedwork}

		\subsection{Page cache}

			In Linux, page cache is a part of memory, which includes pages referring to 
			physical pages on disk \cite{linuxdev3rd2010}. 
			The goal of the page cache is to avoid disk I/O by keeping data in memory 
			for later accesses, minimize I/O time since disk accesses are considerably 
			slower than memory accesses.
			When the kernel starts a read operation, it checks if required data is in page cache.
			If yes, data is read from memory. This is called a \textit{cache hit}. 
			Otherwise, data is read from disk and placed in page cache.
			For writing, page cache is implemented with three different strategies. 
			The first is \textit{direct~I/O} or \textit{no-write}, which means the page cache 
			is not involved in write operations. 
			The second is \textit{write-through}, in which data goes \textit{through} 
			page cache to disk is. In this case, data is written to both page cache and disk. 
			In the third strategy, which is \textit{write-back}, data are written directly 
			to disk as \textit{dirty} pages and these pages will be periodically flushed 
			to disk later by a \textit{writeback} process and then marked as not dirty. 
			The write-back strategy is considered to outperform write-through as well as 
			direct I/O since it delays disk writes and perform in batch later 
			\cite{linuxdev3rd2010}.

			In write-back, one of the keys is the cache eviction mechanism, which 
			removes pages from the page cache when available memory is insufficient. 
			The kernel chooses clean pages from page cache to evict. 
			If the clean pages are inadequate, dirty pages are flushed to disk, 
			marked as clean and available for eviction. 
			Different cache eviction algorithms have been developed and proposed
			\cite{owda2014comparison}.
			In Linux kernel, a \textit{two lists strategy (LRU/2)} is implemented based on 
			the least-recently-used (LRU) scheme, with an \textit{active list} only containing 
			pages accesses more than once, and an \textit{inactive list} containing 
			pages available for eviction. 
			Pages are moved from inactive list to active list if they are 
			accessed when residing in the inactive list. 
			The lists work on LRU manner, pages are added to the tail and removed 
			from the top. 
			The lists are also kept balanced by moving pages from the active list
			to the inactive list when the active list grows too large.

		\subsection{Simulators}

			There has been the research and development of many simulation tools 
			and frameworks for cloud computing with the goal to simulate cloud, grid, 
			HPC platforms, etc. 
			A number of the most famous existing simulators have been analyzed 
			and compared in some reviews (e.g., \cite{byrne2017review}, \cite{casanova2014simgrid}). 
			These simulators provide simulation models of CPU, network, storage 
			with different approaches and levels of details.
			Some simulators are at packet-level, which allows to achieve 
			high simulation accuracy with fine-grain details such as storage simulation 
			in \textit{DiskSim} or network simulation in \textit{CloudSim} 
			and \textit{iCanCloud} \cite{casanova2014simgrid}. 
			However, the issues with the packet-level simulation are scalability due to 
			long simulation time when the simulated platforms and applications grow 
			to hundreds and thousands nodes. 
			SimGrid is an open source simulation platform that has been developed 
			not only to balance the trade-off between accuracy and scalability, 
			but also to achieve versatility, which allows the use of the framework 
			to simulate applications in multiple domains \cite{casanova2014simgrid}.

			However, I/O with page cache model has not been provided in SimGrid. 
			It only simulates storage with capacity and bandwidths.
			Lebre \textit{et al.} adds a storage simulation extension to SimGrid 
			to simulate file system,  but the page cache is not taken into account 
			\cite{lebre2015}. 
			SIMCAN is the simulation framework that models I/O with page cache 
			with data on disk can be kept in page cache \cite{nunez2012simcan}.
			Memory is also modeled in iCanCloud simulator with a component in charge 
			of managing memory accesses and cached data \cite{nunez2012icancloud}, 
			but one of the issues is scalability due to its sophisticated model.
			Ultimately, non of these simulators provide any cache write strategy, 
			cache eviction mechanism with dirty/clean data and the page cache LRU lists. 
			Xu \textit{et al.} applies four cache replacement policies to simulate in-memory 
			caching, but for the study in energy consumption of multi-tier heterogenous 
			network \cite{xu2018saving}.

			In this study, we chose SimGrid to implement our I/O with page cache model
			due to the balancing between accuracy and scalibility, the versatility allowing it 
			to be adopted in different domains, and the design enabling this 
			framework to be extended with new models. 
			The main issue users have to face when using SimGrid is everything has to be 
			implemented from scratch due to its low-level API. 
			To ease this challenge, WRENCH, an open-source library, is built on top 
			of SimGrid to abstract the low-level logic and expose a higher abstraction 
			level API \cite{casanova2018wrench}. 
			We integrate our model in WRENCH and use the SimGrid API to interact with 
			simulated storage resource.
			This approach allows us to leverage existing features in WRENCH and SimGrid.
			The first is the process control, synchronization of resource consumption by 
			concurrent processes handled by the simulation core in SimGrid. 
			The second is the existing SimGrid storage model with capacity and 
			bandwidth properties, which are similar to the characteristics of memory.
			The third is dropping the burden of implementation with the high level 
			of abstraction in WRENCH.

	\section{Method}
	\label{method}

@@

			We leverage existing storage simulation models to simulate the memory,
			which is characterized with storage capacity, distinguished read/write 
			bandwidths and device latency. 
			All of these features are dealt within models in simulation 
			frameworks, which makes storage models applicable to memory \cite{lebre2015}. 
			In addition, leveraging existing disk models allows us 
			to take advantage of bandwidth sharing between concurrent disk I/O to 
			simulate parallel memory accesses.

@@

			When simulated applications need to read/write a file, they can simply call 
			the IO Controller instead of the storage simulation models.

			As mentioned, in our model, a file read/write can be completed by read/write 
			chunk-by-chunk until entire file is read/written.
			To read or write a chunk of file, a simulated application sends a request to the 
			IOController instance on the node where the requested file is stored.
			Based on the memory status in Memory Manager, the IOController 
			orchestrates cache and disk reads/writes, flushing, cache eviction 
			to simulate I/O time.

@@

			We use our dedicated cluster hosted at Concordia University to conduct 
			the experiments. The cluster consists of one login node, 8 compute nodes 
			and 4 storage nodes connected with a network bandwidth of 3GBps. 
			The login node has one Intel(R) Xeon(R) Gold 6130 CPU @ 2.10GHz, 
			137~GB (128~GiB) of RAM, 1.8TB of storage of XFS file system, 
			13~GB of tmpfs file system and 385 TB of Lustre shared file system. 
			Each compute node has two 16 cores Intel(R) Xeon(R) Gold 6130 CPU @ 2.10GHz, 
			275~GB (256~GiB) of RAM and 6 SSDs, 450~GB each with XFS file system, 
			378~GB of tmpfs and 126~GB of devtmpfs file system.
			Lustre file system is configured with one metadata target with 854~GB 
			of storage, 44 object storage targets with 8.7~TB of storage each. 
			The cluster is run on CentOS 8.1 with NFS version 4. 
			We use \textbf{\textit{atop}} and \textbf{\textit{collectl}} as tools to 
			monitor and collect memory status and disk throughput 
			on the cluster.

@@

			\subsubsection{Evaluation on NFS}

		    Since network file systems are very common in HPC, we conduct 
		    the next experiment with the same scenario as in the multi-threaded experiment, 
		    except that files are stored on a node, read and written by another node. 
		    The compute node is the node used in the multi-threaded experiment, 
		    and the capacity of the disk used on the storage node is 400 GB.

		    In NFS, client cache and server cache are enabled in reading, but only 
		    server write-through cache is involved in writing. 
		    Although write-through does not have any impacts writing performance, 
		    caching data on server side allows server to avoid disk reads, which are 
		    significantly longer than cache reads.
		    Thus, we simulate page cache for reading on both client and server node.
		    In writing, write-though is simulated with data being written directly to 
		    server storage and then added to server cache.

@@

	\section{Results}
	\label{results}			
		The benchmark results of bandwidths and the values used in the simulators 
		in our all experiments are shown in \ref{table:benchmark}.

		\begin{table}
		\centering
		\begin{tabular}{|l|c|c|c|}
		\hline
			Bandwidths (MBps)  & Cluster (real) & Python simulator & WRENCH simulator\\
		\hline
			Memory read  & 6860	& 4812	 & 4812\\
			Memory write & 2764	& 4812 & 4812\\
			Local disk read & 510 & 465 & 465\\
			Local disk write & 420 & 465	 & 465\\
			Remote disk read & 515 & - & 445\\
			Remote disk write & 375 & - & 445\\
			Network bandwidth & 3000 & - & 3000\\
		\hline
		\end{tabular}
		\caption{Bandwidth benchmark results and simulator configurations}
		\label{table:benchmark}
		\end{table}

@@

			The I/O time of tasks in simulators and real execution is also tracked and compared. 
			Figure~\ref{fig:single_errors} shows the relative error of simulated I/O time 
			of individual I/O activities. 
			Our pipeline consists of 3 tasks, and each task reads an input file, 
			performs some computation, and writes output file. 
			Thus, there are 6 sequential I/O activities, \textit{Read 1} and \textit{Write 1} 
			from the first task, \textit{Read 2} and \textit{Write 2} the second task, 
			and \textit{Read 3}, \textit{Write3} from the last task.

			In general, with all input sizes in the experiment, the simulation errors 
			of simulators with page cache model (Python and WRENCH with page cache simulators) 
			are remarkably lower than the result with original WRENCH . 
			The reason is, in original WRENCH, because page cache is not simulated, 
			data is only read/written with disk bandwidth, which is much slower 
			than memory bandwidths, leading to simulated I/O time longer than in reality.
			Especially, the simulation time errors of \textit{Read 3} in the simulators 
			with page cache are just under the error in original WRENCH.

@@

			\begin{figure*}		
			\begin{subfigure}{\linewidth}
				\centering
   				\includegraphics[width=\linewidth]{result/multi/figures/multi_local.pdf}
			\end{subfigure}		
			\caption{I/O time of concurrent pipelines with local storage}
			\label{fig:multi_local}
			\end{figure*}

@@

			\begin{figure*}		
			\begin{subfigure}{\linewidth}
				\centering
   				\includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
			\end{subfigure}		
			\caption{I/O time of concurrent pipelines with NFS}
			\label{fig:multi_nfs}
			\end{figure*}

			Similarly, the results with from the experiment with NFS are described 
			by the average I/O time shown in Figure~\ref{fig:multi_nfs}.
			In general, the average I/O time reflect a very similar level of accuracy 
			to the previous experiment results when the I/O time of pipelines 
			is very close to the real execution. 
			The simulated read time if a bit off when the number of pipelines 
			surpasses 22, with 264 GB of data in total. 
			This is due to the impact of simulated cache eviction when 
			the page cache is saturated, which has been seen in 
			Figure~\ref{fig:single_error} with 100 GB of input. 
			When it comes to write time, original WRENCH and WRENCH with page cache 
			results are identical when both are writing directly to disk.

@@

	\section{Discussion and Future Work}
	\label{discussion}			
			In computing infrastructures, especially those use for data-intensive applications, 
			page cache undoubtedly has certain impacts on performance. 
			Conducting simulation experiments is an effective approach to predict, 
			evaluate the performance of not only application, scheduler but also 
			infrastructure settings. 
			Unfortunately, page cache is missing in many simulation tools for HPC systems. 
			In this study, we proposed a simulation model of I/O with page cache, 
			implement it with SimGrid using WRENCH, and evaluate the results 
			with different experiments. 
			The results shows our achievements in terms of simulation accuracy 
			and the application potential in practice. 
			On the other hands, the model exposed some limitations that can be 
			further improved.

			Our short term future work will be re-evaluate the model with asymmetrical 
			disk bandwidths that will be released in the next SimGrid version.
			The next step is to study the detail in cache eviction mechanism to achieve 
			better simulation results. 
			We also plan to support random I/O and file readahead, as well as 
			saving and restoring disk state when hosts are turned on/off.

\bibliographystyle{plain}
\bibliography{citation}
